---
title: "DT Lab"
author: "Ruth Efrem"
date: "11/10/21"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries
```{r}
library(rio)
library(plyr)
library(tidyverse)
library(rpart)
library(psych)
library(pROC)
#install.packages("rpart.plot")
library(rpart.plot)
#install.packages("rattle")
library(rattle)
library(caret)
library(C50) #Need this to pass into caret 
library(mlbench)
```

Congrats! You just graduated UVA's MSDS program and got a job working at the 
Treasury Department. In partnership with Congress members the Treasury has been 
ask to come up with innovative ways to create tax policy. In doing so they 
want to be able to reliably predict whether American workers make more than 
$50,000 a year and also which variables seem to be most contributing 
to predicting this outcome. 

You would like to be able to explain the model to the mere mortals 
around you but need a fairly robust and flexible approach so you've 
chosen to use decision trees to get started and will possibly move 
to a ensemble model if needed. 

In doing so, similar to  great data scientists of the past 
you remembered the excellent education provided 
to you at UVA in a undergrad data science course and have outline 
20ish steps that will need to be undertaken to complete this task 
(you can add more or combine if needed).  As always, you will need 
to make sure to #comment your work heavily. 


```{r}
#url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data"
#xx <- readr::read_csv(url)
#View(xx)
```

 Footnotes: 
-	You can add or combine steps if needed
-	Also, remember to try several methods during evaluation and always be 
mindful of how the model will be used in practice.
- Make sure all your variables are the correct type (factor, character, etc.)


```{r}
#1 Load the data, check for missing data and ensure the labels are correct. 

url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data"
xx <- readr::read_csv(url)
View(xx)

xx <- readr::read_csv(url, col_names = FALSE)

names <- c("age","workclass","fnlwgt","education","education-num","marital-status","occupation","relationship","race","sex","capital-gain","capital-loss","hours-per-week","native-country", "salary")

names(xx) <- names

View(xx)
```

```{r}
#2 Ensure all the variables are classified correctly including the target 
# variable
str(xx)
xx$salary = as.factor(xx$salary)
xx$workclass = as.factor(xx$workclass)
xx$sex = as.factor(xx$sex)
xx$race = as.factor(xx$race)
xx$`native-country`= as.factor(xx$`native-country`)
xx$relationship= as.factor(xx$relationship)
xx$`marital-status`= as.factor(xx$`marital-status`)
xx$education = as.factor(xx$education)
xx = xx[-c(2,4,7,14)]
View(xx)
str(xx)


table(xx$salary)
# data is very imbalanced, much more making greater than or equal to 50,000 dollars than people making less than 50,000 dollars

```

```{r}
#3 Don't check for correlated variables....because it doesn't matter with 
# Decision Trees...the make local greedy decisions. 
```

```{r}
#4 Guess what, you also don't need to standardize the data, 
#because DTs don't give a ish, they make local decisions...keeps getting easier
```

```{r}
#5 Determine the baserate or prevalence for the classifier, 
# what does this number mean?
```

According to the prevalence given further into this code in the confusion matrix, the prevalence is 0.76. Prevalence gives us an indication of the balance of the data set. This means that 76 percent of people are making less than 50,000 dollars a year and 24 percent are making more than or equal to 50,000 dollars per year.

```{r}
#6 Split your data into test, tune, and train. (70/15/15)
#There is not a easy way to create 3 partitions using the createDataPartitions
#so we are going to use it twice. Mostly because we want to stratify on the variable we are working to predict. What does that mean?  
part_index_1 <- caret::createDataPartition(xx$salary,
                                           times=1,
                                           p = 0.70,
                                           groups=1,
                                           list=FALSE)
train <- xx[part_index_1, ]
tune_and_test <- xx[-part_index_1, ]
#The we need to use the function again to create the tuning set 
tune_and_test_index <- createDataPartition(tune_and_test$salary,
                                           p = .5,
                                           list = FALSE,
                                           times = 1)
tune <- tune_and_test[tune_and_test_index, ]
test <- tune_and_test[-tune_and_test_index, ]
dim(train)
dim(test) 
dim(tune)
```

```{r}
#7 Build your model using the training data and default settings in caret, 
# double check to make sure you are using a cross-validation training approach

# Choose the features and classes
features <- train[,c(-2,-8:-11)]#dropping these columns which includes target variable, salary

target <- train$salary
str(features)
str(target)
#Cross validation process 
fitControl <- trainControl(method = "repeatedcv",
                          number = 10,
                          repeats = 5, 
                          returnResamp="all",
                          #classProbs = TRUE,
                          allowParallel = TRUE) 
# number - number of folds
# repeats - number of times the CV is repeated, here it's 5 take the average of
# those 5 repeats
# Grid search options for each of the models available in CARET
# http://topepo.github.io/caret/train-models-by-tag.html#tree-based-model
grid <- expand.grid(.winnow = c(TRUE,FALSE), 
                    .trials=c(1,5,10,15,20), 
                    .model="tree")
#expand.grid - series of options that are available for model training
#winnow - whether to reduce the feature space -  Works to remove unimportant 
#features but it doesn't always work, in the above we are winnowing (process of eliminating variable based off their contribution to the model.  
#Actually a pretty good StackExchange post on winnowing:
#https://stats.stackexchange.com/questions/83913/understanding-the-output-of-c5-0-classification-model-using-the-caret-package
#trails - number of boosting iterations to try, 1 indicates a single model 
#model - type of ml model
set.seed(1984)
xx_mdl <- train(x=features,
                y=target,
                method="C5.0",
                tuneGrid=grid,
                trControl=fitControl,
                verbose=TRUE)

xx_mdl #provides us the hyper-parameters that were selected through the grid



```

```{r}
#8 View the results, what is the most important variable for the tree?

varImp(xx_mdl)

```

The variables : education-num, marital-status,age,capital-gain, and capital-loss are the best ones to use

```{r}
#9 Plot the output of the model to see the tree visually 
set.seed(1984)
xx_mdl <- train(x=features,
                y=target,
                method="C5.0",
                tuneGrid=grid,
                trControl=fitControl,
                verbose=TRUE)
xx_mdl #provides us the hyper-parameters that were selected through the grid
# search process. 
View(xx_mdl$pred)
# visualize the re-sample distributions
xyplot(xx_mdl,type = c("g", "p", "smooth"))
```


```{r}
#10 Use the validation set and the predict function with your model to the
# estimate the target variable.
```

```{r}
#11 Compare the predicted values to those of the actual by generating a 
# matrix ("by-hand").

xx_eval <-(predict(xx_mdl,newdata = test))#generates 2 salary categories
View(xx_eval)

xx_eval_prob <- predict(xx_mdl, newdata = test, type = "prob")#this gives us the predicted prob, we will need these later for the fairness evaluation
View(xx_eval_prob)

View(test$salary)
table(xx_eval, test$salary)

xx_eval_prob$test <- test$salary

View(xx_eval_prob)

(error = mean(xx_eval != test$salary))




# is this "by hand" ?---> 87%
```


```{r}
#12 Use the the confusion matrix function to check a variety of metrics 
# and comment on the metric that might be best for this type of analysis given your question.  

#xx_pred_tune = predict(xx_mdl,tune, type= "raw")
#View(as_tibble(xx_pred_tune))
#Lets use the confusion matrix
(xx_eval <- confusionMatrix(as.factor(xx_pred_tune), 
                as.factor(tune$salary), 
                dnn=c("Prediction", "Actual"), 
                mode = "sens_spec"))
table(tune$salary)
(xx_pred_tune_p = predict(xx_mdl,tune,type= "prob"))

```

The metric that might be best for this type of analysis is...

-predicted positive class rate (precision)
-specificity, unbalanced : use F1
-never bad to look at log loss

```{r}
#13 Generate a ROC and AUC output, interpret the results

library(ROCR)

xx_pred_tune = predict(xx_mdl,tune, type= "raw")
View(as_tibble(xx_pred_tune))

table(tune$salary)
(xx_pred_tune_p = predict(xx_mdl,tune,type= "prob"))
View(xx_pred_tune_p)

xx_eval <- tibble(pred_class= xx_pred_tune, pred_prob=xx_pred_tune_p$`1`,target=as.numeric(test$salary))

View(xx_eval)


```


##############
```{r}
#Pulling out the probabilities  
xx_prob <- tibble(attr(xx_mdl, "prob"))

#Prob that are mixed
View(xx_prob)

#Building a dataframe includes the columns 
final_model <- tibble(k_prob=titanic_prob$`attr(titanic_3NN, "prob")`,pred=titanic_3NN,target=tune$`survived`)

View(final_model)

#Need to convert this to the likelihood to be in the poss class.
pos_prec <- ifelse(final_model$pred == 0, 1-final_model$k_prob, final_model$k_prob)
```
##################



###                         IGNORE FOR NOW
```{r}
#14 Use the predict function to generate percentages, then select several 
# different threshold levels using the confusion matrix function and 
# interpret the results. What patterns did you notice, did the evaluation metrics change? 

adjust_thres <- function(x, y, z) {
  #x=pred_probablities, y=threshold, z=test_outcome
  thres <- as.factor(ifelse(x > y, 1,0))
  confusionMatrix(thres, z, positive = "1" , dnn=c("Prediction", "Actual"), mode = "everything")
}

str(test$salary)

adjust_thres(xx_eval_prob$`>50K`, .60, test$salary) #Not much changes here because of the high probability splits of the data outcomes. Let's take a closer look. We can see that the algo isn't marginally mis-classifying these rows, it's very confidently wrong. Which likely means that there's too much emphasis being placed on too small a number of variables, principally the funfetti variable. 

xx_eval_prob$test <- test$salary

View(xx_eval_prob)

(error = mean(loan_eval != test$salary))#overall error rate, on average when does our prediction not match the actual, looks like around 15%, really just ok. 





```


```{r}
#15 Based on your understanding of the model and data adjust several of the hyper-parameters via the built in train control function in caret or build and try new features, does the model quality improve? If so how and why, if not, why not?
# Use this link: https://rdrr.io/cran/caret/man/trainControl.html to select changes,
# you aren't expected to understand all these options but explore one or two and 
# see what happens. 


```


```{r}
#16 Once you are confident that your model is not improving, via changes 
# implemented on the training set and evaluated on the the validation set (item 16), predict with the test set and report a final evaluation of the model. Discuss the output in comparison with the previous evaluations. 

wine_pred_test = predict(wine_mdl_2,test, type= "raw")
View(as_tibble(wine_pred_test))
#Lets use the confusion matrix
confusionMatrix(as.factor(wine_pred_test), 
                as.factor(test$text_rank), 
                dnn=c("Prediction", "Actual"), 
                mode = "sens_spec")
(wine_pred_tune_p = predict(wine_mdl,test,type= "prob"))


```

```{r}
#17 Summarize what you learned along the way and make recommendations on how this could be used moving forward, being careful not to over promise. 
```

```{r}
#18 What was the most interesting or hardest part of this process and what questions do you still have? 
```

